Master Thesis

### 2013-06-15 the-last-part.tex
Finished drafting Graphical ACGs

\section{Negation}

@@

\section{Agreement}

@@

\section{Intersections of Languages}

@@

We can confirm that the above graphical model of defining ACGs is
coherent. First, we can show that our interpretation is sound and so
that any directed tree adorned with signatures on the nodes, compatible
lexicons on the edges and a distinguished type on the root can only
define ACG languages. This is trivial, since the language corresponding
to the root is an ACG abstract language and the class of ACG languages
is closed under transformation by a lexicon and therefore all of the
languages corresponding to the descendants of the root are ACG languages
as well. Second, we can try and show some kind of completeness. However,
there is not much formalization done to describe a ``system of ACGs that
share common signatures'' (something that our generalization will
fix). We can nevertheless show that it is possible to express any ACG
using our graphical model. Given a grammar $\mathcal{G} =
\mathopen{<}\Sigma_A, \Sigma_O, \mathcal{L}, S\mathclose{>}$, we just
produce a tree whose root has the signature $\Sigma_A$ and distinguished
type $S$ and which is connected to a single child node having the
signature $\Sigma_O$ by an edge labelled by the lexicon
$\mathcal{L}$. Then the two nodes of our tree define both the abstract
and the object language defined by $\mathcal{G}$.

@@

%% \subsubsection{On Alternative Interpretations of Graphical ACGs}
%
% Mention the example with the levels of linguistic representations
% linked by many-to-many transductions and explain how a different
% theory of graphs defining languages might be suitable to that
% architecture.
%
% Also point out the weirdness of the diamond shape.

>>>>

\section{Negation}
\label{sec:negation}

@@

\section{Agreement}
\label{sec:agreement}

@@

\section{Intersections of Languages}
\label{sec:sects-of-langs}

@@

We can confirm that the above graphical model of defining ACGs is
coherent. First, we can show that our interpretation is sound and so
that any directed tree adorned with signatures on the nodes, compatible
lexicons on the edges and a distinguished type on the root can only
define ACG languages. This is trivial, since the language corresponding
to the root is an ACG abstract language and the class of ACG languages
is closed under transformation by a lexicon and therefore all of the
languages corresponding to the descendants of the root are ACG languages
as well. Second, we can try and show some kind of completeness. However,
there is not much formalization done to describe a ``system of ACGs that
share common signatures'' (something that our generalization will
fix). We can nevertheless show that it is possible to express any ACG
using our graphical model. Given a grammar $\mathcal{G} =
\mathopen{<}\Sigma_A, \Sigma_O, \mathcal{L}, S\mathclose{>}$, we just
produce a tree whose root has the signature $\Sigma_A$ and distinguished
type $S$ and which is connected to a single child node having the
signature $\Sigma_O$ by an edge labelled by the lexicon
$\mathcal{L}$. Then the two nodes of our tree define both the abstract
and the object language defined by $\mathcal{G}$.
TODO: Prove here that the modes and patterns of composition in ACG have
the predictable semantics as well.

@@

Having covered some of the formal properties of graphical ACGs, we will
get back to building grammars. Armed with this formalism, we can solve
the puzzle of incorporating the three constraints of sections
\ref{sec:negation}, \ref{sec:extraction} and \ref{sec:agreement}. Our
graph $G$ will consist of 6 nodes, $V(G) = \{Neg, Ext, Agr, Synt, Str,
Sem\}$ with edges going from $Neg$, $Ext$ and $Agr$ to $Synt$ and from
$Synt$ to $Str$ and $Sem$ (see Figure~TODO). The signatures for the
first three named vertices are exactly those that have been presented in
their dedicated sections and it is the case also for the distinguished
types. The lexicons for the edges connecting these constraint nodes to
$Synt$ all behave according to the scheme below:

\begin{align*}
  \mathcal{L}_{(constraint,Synt)}(X_{wordform_{subscript}}) &= C_{wordform} \\
  \mathcal{L}_{(constraint,Synt)}(TYPE\_FEATURES)     &= TYPE
\end{align*}

The $Synt$-$Str$-$Sem$ subgraph is identical to the pair of ACGs
displayed in \ref{sec:acg} except for some additions that need to be
made so that the fragment covers the wordforms that we introduced when
discussing the constraints (prepositions, relative pronouns, the
negative particle $ne$\ldots). However, their types will not be tied to
any of the constraints.

\begin{align*}
  C_{de} &: NP \limp N \limp N \\
  C_{qui} &: (NP \limp S) \limp N \limp N \\
  C_{dit\ que} &: NP \limp S \limp S \\
  C_{dorment} &: NP \limp S \\
  C_{ne_{tv}} &: (NP \limp NP \limp S) \limp (NP \limp NP \limp S) \\
  C_{ne_{iv}} &: (NP \limp S) \limp (NP \limp S) \\
  C_{ne_{cv}} &: (NP \limp S \limp S) \limp (NP \limp S \limp S)
\end{align*}

This lets us describe the high-level combinatorics of language that are
of interest to the categorial grammarian in a manner which is idiomatic
in academic literature (simple focused types $N$, $NP$ and $S$). At the
same time, we can actually use this textbook grammar as a basis for a
more realistic grammar which handles constraints such as agreement and
extraction islands. Furthermore, these additional constraints are also
defined in a manner that has its precedent in existing research. When
the authors of \cite{pogodalla2012controlling} decided to put the
constraints of extraction under a microscope and try to encode them in
ACGs, they defined their constraints in the same way as our architecture
does. This graphical architecture lets us write the syntactic kernel and
the additional constraints independently and using two different styles
that have both been shown useful by their own merits, which we see as
indicative of enabling modularity.

In the above approach, we have defined a language by providing a grammar
which provides us with useful structures but which licences
incorrect/ungrammatical structures. We have then refined this grammar by
tacking on constraints ``on the side''. This approach is generally
useful and can be seen e.g. in \cite{van2004concepts} to define the
formal programming language kernel on which a formal semantics is
established. In our case, we are interested in having intelligible
syntactic structures too as we would like to compute their semantic
representations. The syntax-semantics interface already presents a
significant challenge and having less-than-ideal syntactic structures to
start with does not make formalizing it any easier.

We can imagine a scenario in which graphical ACGs become indispensable
to have a clean syntax-semantics interface. In
\ref{sec:sects-of-langs}(TODO: put here a more specific label after
splitting up the section into subsections), we have seen a pattern of
composing ACGs wherein ambiguities which are of a semantic nature are
moved to a more abstract signature (i.e. the syntactic terms are
transduced to the semantic terms). But there is also another pattern,
that of adding a new abstract signature to constrain the syntactic
terms. Since in classical ACGs, one does not define languages by
providing more than one abstract signature, the two abstract signatures
would have to be stacked on top of each other.

Consider the case when we first add the abstract signature to constrain
our syntactic terms. If we then try to transduce the terms of this new
abstract signature, we will have to deal with the whole medley of types
that were introduced in the constraining signature. This problem occurs
since in our type-logical grammar the type system of the syntactic
signature does not provide just the interface of the syntactic layer,
but also its implementation. By putting the constraining signature in a
different subgraph, aside from the syntax-semantics transducer, we can
shield the syntax-semantics interface from the gnarly implementation
details of the rules of syntax. This in turn gives us an answer as to
what level of information should be expressed in the syntactic signature
itself and which other constraints should be tucked away in hidden
abstract signatures: the syntactic signature to which the
syntax-semantics transducer is connected should only contain information
relevant to the syntax-semantics interface. (TODO: Think about this some
more, about whether there is any important difference to the
syntax-semantics interface when the syntactic signature does not enforce
anything, e.g..)

As for the other case where we first build the syntax-semantics
transducer and than constrain its abstract signature with a more
abstract one is not much better. Now we have a converse problem where
the implementation of the syntactic constraints is complected with bits
of semantics since the constraining signature will have to map onto a
signature built around types combining both syntax and semantics. (TODO:
Go to bed.)

% We could have talked here about making the graph of constraints deeper
% to ``reuse'' common concepts and types. However due to the need for
% more abstract signatures to enforce and basically reimplement the
% constraints in less abstract signatures this would not be so
% useful.

% We could also talk about how constraints could be added to semantic or
% morphological signatures, but we have nothing interesting to say about
% that.

\subsubsection{On Alternative Interpretations of Graphical ACGs}

Our definitions of intrinsic and extrinsic have consequences which might
be counter-intuitive and not correspond to the grammarian's intent. We
will describe what these consequences are and propose an alternative
notion of a language defined by a graphical ACG which aims to reconcile
them.

The first scenario which can be tricky is when two paths from nodes
$u_1$ and $u_2$ converge on a single node $v$. If we then take a term
$\alpha$ from $\mathcal{E}_{\mathcal{G}}(u_1)$, the term
$\mathcal{L}_{(u_1,v)}(\alpha)$ belongs to
$\mathcal{I}_{\mathcal{G}}(v)$ but not necessarily to
$\mathcal{E}_{\mathcal{G}}(v)$ since it might be the case that there is
no antecedent for $\mathcal{L}_{(u_1,v)}(\alpha)$ in
$\mathcal{E}_{\mathcal{G}}(u_2)$. To sum up, mapping a member of some
$\mathcal{E}_{\mathcal{G}}(u)$ using some $\mathcal{L}_{(u,v)}$ does not
guarantee that the result belongs to $\mathcal{E}_{\mathcal{G}}(v)$,
which is contrary to the case of object languages in ACGs.

This highlights some limitations of our notion of an extrinsic
language. Consider the graphical ACG of Figure~TODO. We have more than
one path converging to $Synt$. We will consider a term $\alpha$ of
$\mathcal{E}_{\mathcal{G}}(Sem)$. $\alpha$ has an antecedent in
$\mathcal{E}_{\mathcal{G}}(SyntSem)$ which has a descendant $\beta$ in
$\mathcal{I}_{\mathcal{G}}(Synt)$. However, because of the constraints
on $Synt$ ($Neg$, $Ext$ and $Agr$), $\beta$ can be syntactic term
violating one of these constraints which yields an ungrammatical
sentence. Therefore, the notion of an extrinsic language is insufficient
to express the set of meaning terms from $\Lambda(\Sigma_{Sem})$ which
are expressible by grammatical sentences. Such definitions would be
useful in an example like the one in Figure~TODO where we have a chain
of signatures representing the different levels of linguistic
description whose terms are in many-to-many relationships expressed
through transducers. In such a scenario, it would be desirable to be
able to express the set of descriptions belonging to one level,
e.g. morphology, that are related to valid descriptions on all the other
levels.

Compared to ACG object languages, we lose another guarantee. In ACGs,
the central role of the root abstract-most signature guarantees us that
whenever a term belongs to an object language, we can find terms in all
the other object languages that all share the same abstract-most
term. On the other hand in the intrinsic and extrinsic languages we
defined on graphical ACGs, having a term in an extrinsic language of
some node $v$ only guarantees the presence of corresponding terms in the
extrinsic languages of its ancestors and the intrinsic languages of
their descendants.

The second puzzling scenario is the diamond shape on Figure~TODO. If we
consider a term $\delta$ of $\mathcal{E}_{\mathcal{G}}(Bottom)$, we know
that there exist antecedents $\beta$ and $\gamma$ in
$\mathcal{E}_{\mathcal{G}}(Left)$ and
$\mathcal{E}_{\mathcal{G}}(Right)$. And while there definitely exist
antecedents $\alpha$ and $\alpha'$ to both $\beta$ and $\gamma$ in
$\mathcal{E}_{\mathcal{G}}(Top)$, these antecedents can be different
disjoint, meaning there does not have to be a single $\alpha \in
\mathcal{E}_{\mathcal{G}}(Top)$ which yields $\beta \in
\mathcal{E}_{\mathcal{G}}(Left)$ and $\gamma \in
\mathcal{E}_{\mathcal{G}}(Right)$ such that they both yield
$\delta$. This means that the result of parsing a term in some
$\mathcal{E}_{\mathcal{G}}(v)$ is a collection of abstract terms with
one term not just for every node $u$ that is more abstract than $v$ but
for every path leading to $v$. Parsing in our diamond grammar would thus
produce four terms $\beta$, $\gamma$, $\alpha$ and $\alpha'$
corresponding to the four paths $(Left,Bottom)$, $(Right,Bottom)$,
$(Top,Left,Bottom)$ and $(Top,Right,Bottom)$, respectively. We contrast
this again with the case of ACGs where the result of parsing can be
represented by associating a term to every signature in the
arborescence.\footnote{One could also say that the result of parsing in
  an ACG is just the antecedent in the abstract-most signature and that
  the other terms can then be easily computed by application of
  lexicons. Similarly, we could say that the result of parsing in a
  general graphical ACG is a collection of abstract terms for every
  \emph{maximal} path leading to the node defining the language in
  question. What might be more intuitive, though, is to have some theory
  of graphical ACGs where the result of parsing is a collection of
  antecedents in all the abstract-most nodes.}

We will propose a new way of assigning languages to nodes in a graphical
ACG that will conserve some of the intuitive behavior we are accustomed
to from ACGs. We need not modify the definition of graphical ACGs to do
so since it is only considered with the structure of the graph and its
decoration with ACG paraphernalia, we only define a new kind of
languages.

Let $\mathcal{G} = \mathopen{<} G, \Sigma, S, \mathcal{L} \mathclose{>}$
be a graphical ACG. We define the \emph{pangraphical language} of node
$u$, $\mathcal{P}_{\mathcal{G}}(u)$, to be the set of terms $t$ such
that there exists a labeling of nodes with terms $T$ satisfying the
following conditions:

\begin{itemize}
  \item $T_u = t$.
  \item For all $v \in V(G)$, $\vdash_{\Sigma_v} T_v : S_v$.
  \item For all $(v,w) \in E(G)$, $\mathcal{L}_{(v,w)}(T_v) = T_w$.
\end{itemize}

TODO!: Show that the pangraphical languages solve all the problems
outlined above.

TODO: $\mathcal{P}_{\mathcal{G}}(u) \subset \mathcal{E}_{\mathcal{G}}(u)
\subset \mathcal{I}_{\mathcal{G}}(u)$

TODO: On arborescences, $\mathcal{P}_{\mathcal{G}}(u) =
\mathcal{E}_{\mathcal{G}}(u)$.

TODO?: Show that extrinsic languages are included in pangraphical
languages.

TODO?: Are pangraphical languages also included in extrinsic languages?

TODO?: Write a parsing algorithm for pangraphical languages similar to
the one for extrinsic languages.

\subsubsection{On the Practical Consequences of Graphical ACGs}

TODO: The trivial lexicons we have for our constraints can be
effectively inverted, giving us effective constraint verification. Using
filtering and/or supertagging could also make it efficient.

TODO: Constraint signatures can use specific type extensions, their
added complexity is then contained within the constraint. Furthermore,
we can treat the constraint signature as a black box and in the context
of an implementation, we replace it with a special purpose procedure
which verifies the constraint efficiently.

TODO?: Practical applications: robustness by being able to parse
sentences which violate some constraints, fine-grained syntax checking
(pinpoint the constraint that is being violated).

TODO?: We will need to define a lot of signatures which look exactly
like some other signature but with some small refinements. The ways we
will refine the signature will hopefully not be so diverse and we would
hope to develop some tools to make formulating these refinements
feasible compared to having to restate the entire grammar every time we
add a new constraint.
